---
title: "TEVp FRET Cleavage Assay"
author: "Viktor Jagd Vad"
date: "2026-06-03"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```


# Load packages and data

all the .jws files were removed before hand, so only the exported .csv files were imported. all spectra from 490nm to 620nm was selected

```{r}
library(tidyverse)
library(hms)
library(patchwork)
library(knitr)

# Folder paths
day1_folder <- "/Users/viktorvad/Desktop/Day1_FRET_data"
day6_folder <- "/Users/viktorvad/Desktop/Day6_FRET_data"

# Function for reading JASCO CSV files
read_jasco_csv <- function(file) {
  lines <- readLines(file)
  
  time_line <- lines[str_detect(lines, "^TIME;")]
  time <- str_remove(time_line, "^TIME;")
  time <- time[1]
  
  sample_line <- lines[str_detect(lines, "^Sample name;|^TITLE;")]
  sample_name <- str_remove(sample_line, "^Sample name;|^TITLE;")
  sample_name <- sample_name[1]
  
# Find where fluorescence data starts
xydata_line <- which(str_trim(lines) == "XYDATA")
start <- xydata_line + 1

# Extract fluorescence data
xy_lines <- lines[start:length(lines)]

# Keep only rows starting with numbers
xy_lines <- xy_lines[str_detect(xy_lines, "^\\s*[0-9]")]

# Create table
xy <- tibble(raw = xy_lines)

# Split wavelength and signal columns
xy <- separate(
  xy,
  raw,
  into = c("wavelength", "signal"),
  sep = "\\s+|;",
  extra = "drop",
  fill = "right"
)

# Convert commas to decimal points
xy$wavelength <- as.numeric(str_replace(xy$wavelength, ",", "."))
xy$signal <- as.numeric(str_replace(xy$signal, ",", "."))

# Add metadata
xy$time <- time
xy$sample_name <- sample_name
xy$file <- basename(file)

# Return cleaned dataset
return(xy)
}
```

```{r}
# Import Day 1 data

# Find all CSV files in the Day 1 folder
files_day1 <- list.files(day1_folder, pattern = "\\.csv$", full.names = TRUE)

# Read each CSV file using the read_jasco_csv function
day1_list <- list()

for (i in 1:length(files_day1)) {
  day1_list[[i]] <- read_jasco_csv(files_day1[i])
}

# Combine all Day 1 files into one table
FRET_results_Day1 <- bind_rows(day1_list)

# Keep only wavelengths between 490 and 620 nm
FRET_results_Day1 <- filter(FRET_results_Day1, wavelength >= 490, wavelength <= 620)

# Add a column showing that these data are from Day 1
FRET_results_Day1$day <- "Day 1"
# Import Day 6 data

# Find all CSV files in the Day 6 folder
files_day6 <- list.files(day6_folder, pattern = "\\.csv$", full.names = TRUE)

# Read each CSV file using the read_jasco_csv function
day6_list <- list()

for (i in 1:length(files_day6)) {
  day6_list[[i]] <- read_jasco_csv(files_day6[i])
}

# Combine all Day 6 files into one table
FRET_results_Day6 <- bind_rows(day6_list)

# Keep only wavelengths between 490 and 620 nm
FRET_results_Day6 <- filter(FRET_results_Day6, wavelength >= 490, wavelength <= 620)

# Add a column showing that these data are from Day 6
FRET_results_Day6$day <- "Day 6"
```

## Prepare sample labels

The dilution labels were replaced with the enzyme concentration. 

```{r}
# Replace dilution labels with enzyme concentrations

FRET_results_Day1$sample_name <- str_replace(FRET_results_Day1$sample_name, "1:20", "5 µM")
FRET_results_Day1$sample_name <- str_replace(FRET_results_Day1$sample_name, "1:40", "2.5 µM")
FRET_results_Day1$sample_name <- str_replace(FRET_results_Day1$sample_name, "1:80", "1.25 µM")

FRET_results_Day6$sample_name <- str_replace(FRET_results_Day6$sample_name, "1:20", "5 µM")
FRET_results_Day6$sample_name <- str_replace(FRET_results_Day6$sample_name, "1:40", "2.5 µM")
FRET_results_Day6$sample_name <- str_replace(FRET_results_Day6$sample_name, "1:80", "1.25 µM")
```

## Inspect raw spectra

The full  spectra were first plotted to check whether the data had any outliers or unexpected signal, that needed to be sorted.

```{r}
# Plot Day 1 fluorescence spectra
plot_day1 <- ggplot(
  data = FRET_results_Day1,
  aes(
    x = wavelength,
    y = signal,
    color = sample_name,
    group = file
  )
) +

  # Draw fluorescence curves
  geom_line(linewidth = 0.4) +

  # Add plot labels
  labs(
    title = "Day 1 raw emission spectra",
    x = "Wavelength (nm)",
    y = "Fluorescence signal",
    color = "Sample"
  ) +

  # Use a simple plot theme
  theme_minimal(base_size = 12)


# Plot Day 6 fluorescence spectra
plot_day6 <- ggplot(
  data = FRET_results_Day6,
  aes(
    x = wavelength,
    y = signal,
    color = sample_name,
    group = file
  )
) +

  # Draw fluorescence curves
  geom_line(linewidth = 0.4) +

  # Add plot labels
  labs(
    title = "Day 6 raw emission spectra",
    x = "Wavelength (nm)",
    y = "Fluorescence signal",
    color = "Sample"
  ) +

  # Use a simple plot theme
  theme_minimal(base_size = 12)


# Show both plots together
plot_day1 + plot_day6
```

# N0it5 5 µM spectra over time (day1)

To visualize the spectral development over time, the N0it5 5 µM day 1 sample was selected as a representative example. Duplicate measurements were averaged into 12-minute cycles. so the duplicates which had 30 secunds between reading was averaged. 

```{r}
# Function for preparing N0it5 5 µM fluorescence spectra
make_n0it5_5uM_spectra <- function(data) {
  
  # Keep only the N0it5 5 µM sample
  data <- filter(data, sample_name == "N0it5 5 µM")
  
  # Convert time into time format
  data$time <- hms::as_hms(data$time)
  
  # Sort data by time and wavelength
  data <- arrange(data, time, wavelength)
  
  # Give each measurement time an ID number
  data <- group_by(data, time)
  
  data <- mutate(
    data,
    measurement_id = cur_group_id()
  )
  
  data <- ungroup(data)
  
  # Group duplicate measurements together
  data <- mutate(
    data,
    pair_id = ceiling(measurement_id / 2)
  )
  
  # Average duplicate measurements
  data <- group_by(data, day, pair_id, wavelength)
  
  data <- summarise(
    data,
    signal = mean(signal, na.rm = TRUE),
    .groups = "drop"
  )
  
  # Create time labels
  data <- mutate(
    data,
    time_min = (pair_id - 1) * 12
  )
  
  data <- mutate(
    data,
    time_label = paste0(time_min, " min")
  )
  
  # Order time labels correctly
  data$time_label <- factor(
    data$time_label,
    levels = paste0(sort(unique(data$time_min)), " min")
  )
  
  # Return cleaned dataset
  return(data)
}


# Prepare Day 1 data
day1_n0it5 <- make_n0it5_5uM_spectra(FRET_results_Day1)

# Prepare Day 6 data
day6_n0it5 <- make_n0it5_5uM_spectra(FRET_results_Day6)

# Combine Day 1 and Day 6 data
n0it5_5uM_spectra <- bind_rows(day1_n0it5, day6_n0it5)
```

```{r}
ggplot(
  n0it5_5uM_spectra |>
    filter(
      wavelength >= 500,
      wavelength <= 600,
      day == "Day 1"
    ),
  aes(
    x = wavelength,
    y = signal,
    group = time_label,
    color = time_label
  )
) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Wavelength (nm)",
    y = "Fluorescence intensity (RFU)",
    color = "Time"
  ) +
  theme_minimal(base_size = 23)
```



```{r}
# Select control samples from Day 1
control_spectra_day1 <- filter(
  FRET_results_Day1,
  str_detect(sample_name, "ctrl|control|sub ctrl")
)

# Convert time to time format
control_spectra_day1$time <- hms::as_hms(control_spectra_day1$time)

# Sort by time and wavelength
control_spectra_day1 <- arrange(control_spectra_day1, time, wavelength)

# Give each measurement an ID
control_spectra_day1 <- group_by(control_spectra_day1, time)
control_spectra_day1 <- mutate(control_spectra_day1, measurement_id = cur_group_id())
control_spectra_day1 <- ungroup(control_spectra_day1)

# Group duplicate measurements together
control_spectra_day1$pair_id <- ceiling(control_spectra_day1$measurement_id / 2)

# Average duplicate control measurements
control_spectra_day1 <- group_by(control_spectra_day1, day, pair_id, wavelength)
control_spectra_day1 <- summarise(
  control_spectra_day1,
  signal = mean(signal, na.rm = TRUE),
  .groups = "drop"
)

# Create time labels
control_spectra_day1$time_min <- (control_spectra_day1$pair_id - 1) * 12
control_spectra_day1$time_label <- paste0(control_spectra_day1$time_min, " min")

```

```{r}
# Select NL0it5 Day 1 data between 500 and 600 nm
n0it5_day1_control_subtracted <- filter(
  n0it5_5uM_spectra,
  day == "Day 1",
  wavelength >= 500,
  wavelength <= 600
)

# Add the control signal to the NL0it5 data
n0it5_day1_control_subtracted <- left_join(
  n0it5_day1_control_subtracted,
  select(control_spectra_day1, wavelength, time_label, control_signal = signal),
  by = c("wavelength", "time_label")
)

# Subtract control signal from sample signal
n0it5_day1_control_subtracted$corrected_signal <-
  n0it5_day1_control_subtracted$signal -
  n0it5_day1_control_subtracted$control_signal

```

```{r}
# Plot control-subtracted spectra
ggplot(
  n0it5_day1_control_subtracted,
  aes(
    x = wavelength,
    y = corrected_signal,
    group = time_label,
    color = time_label
  )
) +
  geom_line(linewidth = 1.1) +
  labs(
    x = "Wavelength (nm)",
    y = "Fluorescence intensity (RFU)",
    color = "Time"
  ) +
  theme_minimal(base_size = 23)
```

# Calculate FRET ratio


To calculate the FRET ratio, the donor and acceptor regions were extracted from each spectrum. The donor signal was defined as the average signal from 512–522 nm (F517), and the acceptor signal was  then defined as the average signal from 587–597 nm (F592). The ratio was then calculated like this:

\[
R = \frac{F_{517}}{F_{592}}
\]

```{r}
# Function for selecting donor and acceptor wavelength regions
select_fret_wavelengths <- function(data) {
  
  # Create a new column showing whether the wavelength
  # belongs to the donor or acceptor region
  data <- mutate(
    data,
    wavelength_group = case_when(
      wavelength >= 512 & wavelength <= 522 ~ "F517",
      wavelength >= 587 & wavelength <= 597 ~ "F592",
      TRUE ~ NA_character_
    )
  )
  
  # Remove wavelengths outside the selected regions
  data <- filter(data, !is.na(wavelength_group))
  
  # Group data by day, time, sample, file, and wavelength region
  data <- group_by(
    data,
    day,
    time,
    sample_name,
    file,
    wavelength_group
  )
  
  # Calculate average fluorescence signal
  data <- summarise(
    data,
    signal = mean(signal, na.rm = TRUE),
    .groups = "drop"
  )
  
  # Return cleaned dataset
  return(data)
}


# Select FRET wavelength regions for Day 1
all_data_selected_Day1 <- select_fret_wavelengths(FRET_results_Day1)

# Select FRET wavelength regions for Day 6
all_data_selected_Day6 <- select_fret_wavelengths(FRET_results_Day6)
```

# Correct FRET ratio and calculate fraction cleaved

The raw ratio was corrected using the substrate control. First, the change in the substrate control ratio over time was calculated. This control drift was then subtracted from the enzyme samples to obtain the corrected ratio, which i callet R_corr.

The fraction cleaved was calculated by normalizing R_corr between the first observed corrected ratio (Rmin) and the Rmax was set to 15.

\[
f_{cleaved} = \frac{R_{corr} - R_{min}}{R_{max} - R_{min}}
\]



```{r}
process_fret_data <- function(all_data_selected, Rmax_value = 15) {
  
  # Convert time and calculate time in minutes
  all_data_selected$time_hms <- as_hms(all_data_selected$time)
  
  all_data_selected <- arrange(all_data_selected, day, time_hms)
  
  all_data_selected <- group_by(all_data_selected, day)
  
  all_data_selected <- mutate(
    all_data_selected,
    time_min = as.numeric(time_hms - min(time_hms), units = "mins")
  )
  
  all_data_selected <- ungroup(all_data_selected)
  
  
  # Add sample information
  data_clean <- mutate(
    all_data_selected,
    enzyme = case_when(
      str_detect(sample_name, "L0it4") ~ "L0it4",
      str_detect(sample_name, "N0it5") ~ "N0it5",
      str_detect(sample_name, "sub ctrl") ~ "control",
      TRUE ~ NA_character_
    ),
    concentration = str_extract(sample_name, "\\d+\\.?\\d* µM"),
    condition = case_when(
      sample_name == "sub ctrl" ~ "control",
      TRUE ~ "enzyme"
    )
  )
  
  
  # Put F517 and F592 into separate columns
  data_clean <- pivot_wider(
    data_clean,
    names_from = wavelength_group,
    values_from = signal
  )
  
  
  # Calculate FRET ratio
  data_clean$R <- data_clean$F517 / data_clean$F592
  
  
  # Average duplicate measurements
  data_clean <- arrange(data_clean, day, sample_name, time_min)
  
  data_clean$cycle <- round(data_clean$time_min / 12)
  
  data_avg <- group_by(
    data_clean,
    day,
    sample_name,
    enzyme,
    concentration,
    condition,
    cycle
  )
  
  data_avg <- summarise(
    data_avg,
    time_min = mean(time_min, na.rm = TRUE),
    R = mean(R, na.rm = TRUE),
    n_measurements = n(),
    .groups = "drop"
  )
  
  
  # Calculate control drift
  control_drift <- filter(data_avg, condition == "control")
  
  control_drift <- arrange(control_drift, day, time_min)
  
  control_drift <- group_by(control_drift, day)
  
  control_drift <- mutate(
    control_drift,
    first_R = first(R),
    control_drift = R - first_R
  )
  
  control_drift <- ungroup(control_drift)
  
  control_drift <- select(control_drift, day, cycle, control_drift)
  
  
  # Subtract control drift from all samples
  data_corrected <- left_join(
    data_avg,
    control_drift,
    by = c("day", "cycle")
  )
  
  data_corrected$control_drift <- replace_na(data_corrected$control_drift, 0)
  
  data_corrected$R_corr <- data_corrected$R - data_corrected$control_drift
  
  
  # Calculate fraction cleaved
  data_corrected <- group_by(data_corrected, day, sample_name)
  
  data_corrected <- mutate(
    data_corrected,
    Rmin = R_corr[cycle == min(cycle)][1],
    Rmax = Rmax_value,
    f_cleaved = (R_corr - Rmin) / (Rmax - Rmin),
    f_cleaved = pmin(pmax(f_cleaved, 0), 1)
  )
  
  data_corrected <- ungroup(data_corrected)
  
  
  # Return final processed dataset
  return(data_corrected)
}


# Process Day 1 and Day 6 data
Day1_corr <- process_fret_data(all_data_selected_Day1, Rmax_value = 15)
Day6_corr <- process_fret_data(all_data_selected_Day6, Rmax_value = 15)

# Combine both days
all_corr <- bind_rows(Day1_corr, Day6_corr)
```

```{r}
# ---------- Day 0 ----------
ggplot(
  Day1_corr |> filter(condition == "enzyme"),
  aes(
    x = time_min,
    y = f_cleaved,
    color = concentration,
    group = interaction(enzyme, concentration)
  )
) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ enzyme) +
  labs(
    title = "Day 1",
    x = "Time (min)",
    y = "Fraction cleaved",
    color = "Conc."
  ) +
  theme_minimal(base_size = 14)
# ---------- Day 5 ----------
ggplot(
  Day6_corr |> filter(condition == "enzyme"),
  aes(
    x = time_min,
    y = f_cleaved,
    color = concentration,
    group = interaction(enzyme, concentration)
  )
) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ enzyme) +
  labs(
    title = "Day 6",
    x = "Time (min)",
    y = "Fraction cleaved",
    color = "Conc."
  ) +
  theme_minimal(base_size = 14)
```




## Fraction cleaved

This plot shows the normalized reaction progress. A value of 0 corresponds to the first observed time point, while a value of 1 corresponds to the assumed plateau ratio of 15.

```{r}
n0it5_5uM_compare <- all_corr |>
  filter(
    enzyme == "N0it5",
    concentration == "5 µM",
    condition == "enzyme"
  )


ggplot(
  n0it5_5uM_compare,
  aes(x = time_min, y = f_cleaved, color = day, group = day)
) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.7) +
  labs(
    x = "Time (min)",
    y = "Fraction cleaved",
    color = "Day"
  ) +
  theme_minimal(base_size = 14)
```

# Initial rate estimation

The initial rate was estimated from the first three time points of `f_cleaved`. A linear regression was fitted to these points. The slope has units of fraction cleaved per minute. This slope was multiplied by the substrate concentration, 5 µM, to estimate the initial rate in µM substrate cleaved per minute.

```{r}
initial_rates <- function(data_corr, substrate_conc = 5) {
  
  # Keep only enzyme samples
  data_corr <- filter(data_corr, condition == "enzyme")
  
  # Group data by day, enzyme, and concentration
  data_corr <- group_by(data_corr, day, enzyme, concentration)
  
  # Sort by time
  data_corr <- arrange(data_corr, time_min)
  
  # Keep only the first three time points
  data_corr <- slice_head(data_corr, n = 3)
  
  # Calculate initial rate for each group
  rates <- group_modify(data_corr, ~ {
    
    # Fit a straight line to fraction cleaved vs time
    fit <- lm(f_cleaved ~ time_min, data = .x)
    
    # Extract the slope
    slope <- coef(fit)[["time_min"]]
    
    # Convert slope to initial rate
    initial_rate <- slope * substrate_conc
    
    # Return results
    tibble(
      slope_fraction_per_min = slope,
      initial_rate_uM_per_min = initial_rate,
      r_squared = summary(fit)$r.squared,
      n_points = nrow(.x)
    )
  })
  
  rates <- ungroup(rates)
  
  return(rates)
}


# Calculate initial rates for Day 1
Day1_rates <- initial_rates(Day1_corr, substrate_conc = 5)

# Calculate initial rates for Day 6
Day6_rates <- initial_rates(Day6_corr, substrate_conc = 5)

# Combine Day 1 and Day 6 rates
all_rates <- bind_rows(Day1_rates, Day6_rates)

# Select only N0it5 at 5 µM
n0it5_5uM_rates_compare <- filter(
  all_rates,
  enzyme == "N0it5",
  concentration == "5 µM"
)
```

```{r}
ggplot(
  n0it5_5uM_rates_compare,
  aes(x = day, y = initial_rate_uM_per_min)
) +
  geom_col(width = 0.6) +
  labs(
    title = "N0it5 5 µM: initial rate comparison",
    x = "Experiment",
    y = "Initial rate (µM/min)"
  ) +
  theme_minimal(base_size = 14)
```

```{r}
# Create a table showing Day and initial reaction rate
rate_table <- data.frame(
  Day = n0it5_5uM_rates_compare$day,
  initial_rate_uM_per_min =
    n0it5_5uM_rates_compare$initial_rate_uM_per_min
)

# Show table
rate_table
```

